% Aliveness: Appendices Only
% Elias Kunnas, 2025
% Standalone compilation of all appendices

\input{preamble}

\begin{document}

% ============================================================================
% FRONT MATTER
% ============================================================================
\frontmatter

% Title page
\begin{titlepage}
\centering
\vspace*{2cm}

{\Huge\bfseries Aliveness\par}
\vspace{0.5cm}
{\LARGE Principles of Telic Systems\par}
\vspace{1cm}
{\huge\textbf{Appendices}\par}
\vspace{2cm}
{\Large Elias Kunnas\par}
\vspace{1cm}
{\large October 2025\par}

\vfill

{\small This document contains only the appendices.\par}
{\small For the full book, see the main volume.\par}

\end{titlepage}

% Define acronyms for references
\acrodef{SORT}{Sovereignty-Organization-Reality-Telos Framework}
\acrodef{pSORT}{Personal SORT (individual-level application)}
\acrodef{tSORT}{Tribal SORT (sub-civilizational groups)}
\acrodef{SORTVC}{SORT plus Vitality and Constraint modifiers}
\acrodef{POSIWID}{Purpose Of a System Is What It Does}
\acrodef{IFHS}{Integrity, Fecundity, Harmony, Synergy (The Four Foundational Virtues)}

% Table of Contents
\tableofcontents
\cleardoublepage

% ============================================================================
% APPENDICES
% ============================================================================
\mainmatter

% Don't use \appendix command - configure manually for better control
% Reset counters for appendices
\setcounter{part}{0}
\setcounter{chapter}{0}  % Reset to 0, will become 1 with first \chapter
\renewcommand{\thepart}{\Alph{part}}  % Use letters for part numbers

% Configure appendix numbering: Parts are A/B/C, Appendices are A/B/C...
% Appendices use \subsection directly under \chapter, so format as A.1, A.2, B.1, B.2, etc.
\renewcommand{\thechapter}{\Alph{chapter}}  % Appendices numbered A, B, C...
\renewcommand{\thesection}{\thechapter.\arabic{section}}
\renewcommand{\thesubsection}{\thechapter.\arabic{subsection}}
\renewcommand{\thesubsubsection}{\thesubsection.\arabic{subsubsection}}

% Mark that we're in appendix mode for TOC
\renewcommand{\appendixname}{Appendix}
\renewcommand{\chaptername}{Appendix}

% Configure cleveref to call chapters "Appendix" in the appendix section
\crefname{chapter}{Appendix}{Appendices}

% ============================================================================
% PART A: THE CANONICAL REFERENCE (The Fortress of Gnosis)
% ============================================================================
\part{The Canonical Reference}
\label{part:canonical-reference}
\markboth{Part A: The Canonical Reference}{Part A: The Canonical Reference}

\vspace{-2em}  % Reduce space after part title
\noindent\textit{Core reference material establishing the framework's canonical definitions, empirical foundations, and falsification criteria. The fortress of Gnostic clarity.}
\vspace{1em}

\input{appendix_a_lexicon}
\input{appendix_b_scoring_rubrics}
\input{appendix_c_falsification_protocols}
\input{appendix_d_methodology}
\input{appendix_e_case_studies}

% ============================================================================
% PART B: THE APPLIED PRAXIS (The Armory)
% ============================================================================
\part{The Applied Praxis}
\label{part:applied-praxis}
\markboth{Part B: The Applied Praxis}{Part B: The Applied Praxis}

\vspace{-2em}
\noindent\textit{Tactical and engineering manuals for implementing the framework in the real world. The armory of actionable tools.}
\vspace{1em}

\input{appendix_f_axiological_warfare}
\input{appendix_g_american_refounding}

% ============================================================================
% PART C: THE RESEARCH FRONTIER (The Laboratory)
% ============================================================================
\part{The Research Frontier}
\label{part:research-frontier}
\markboth{Part C: The Research Frontier}{Part C: The Research Frontier}

\vspace{-2em}
\noindent\textit{Speculative extensions, ongoing synthesis, and future research directions. The laboratory of conscious becoming.}
\vspace{1em}

\input{appendix_h_generative_medicine}
\input{appendix_i_influences}
\input{appendix_j_research_program}
\input{appendix_k_ai_alignment}
\input{appendix_l_psychological_frameworks}
\input{appendix_m_gnostic_bestiary}

% ============================================================================
% INDEX
% ============================================================================
\cleardoublepage
\addcontentsline{toc}{chapter}{Index}
\printindex

\end{document}
