\chapter{Falsification Protocols}
\label{app:falsification}

\subsection{\texorpdfstring{\textbf{I. The Integrity Pledge}}{I. The Integrity Pledge}}\label{i.-the-integrity-pledge}

A theory that cannot be falsified is dogma. This appendix lists specific, observable conditions that would shatter this framework. The goal is truth about civilizational dynamics, not framework defense. Evidence-based falsification is victory, not defeat.

\subsection{\texorpdfstring{\textbf{II. Falsifying the Inputs: The SORTVC Axes}}{II. Falsifying the Inputs: The SORTVC Axes}}\label{ii.-falsifying-the-inputs-the-sortvc-axes}

The entire framework rests on the hypothesis that the six SORTVC axes are a necessary and sufficient set of variables to describe a polity's axiological state. The discovery of a polity or phenomenon that cannot be meaningfully mapped by these axes would be a catastrophic failure.

\textbf{1. The SORT Axes Completeness Test:}
\begin{itemize}
\item \textbf{The Hypothesis:} The four SORT axes (S/O/R/T) are necessary and sufficient to describe a polity's axiological state. Any civilization's value configuration can be meaningfully positioned in 4D SORT space.
\item \textbf{The Falsification Condition:} Systematic analysis of diverse historical civilizations reveals that \textbf{SORT scores fail to predict or explain outcomes} that are instead explained by a \textbf{fifth independent axis} not reducible to S/O/R/T combinations. For example: if civilizations cluster into patterns requiring an additional "Harmony/Conflict" dimension orthogonal to all four SORT axes, or if a "Sacred/Profane" axis proves necessary and independent of R (Mythos/Gnosis). Discovery of such a dimension would prove SORT is incomplete.
\item \textbf{Threshold:} Not isolated exceptions, but systematic patterns across \textgreater{}10 civilizations in multiple cultural contexts where SORT + proposed fifth axis significantly outperforms SORT alone in outcome prediction (measurable via regression analysis: ΔR² \textgreater{} 0.15).
\end{itemize}

\textbf{2. The V-Axis Falsification Protocol (Vitality):}
\begin{itemize}
\item \textbf{The Hypothesis:} A civilization's long-term survival and power are a direct function of its Vitality (Fecundity, Productivity, Synergy). A low-V polity is a dying polity.
\item \textbf{The Falsification Condition:} The discovery of \textbf{multiple civilizations (≥3) that sustain low Vitality (V \textless{} 3.0: TFR \textless{} 1.2, GDP stagnation or decline, low trust/Synergy) for \textgreater{}5 generations (\textgreater{}125 years) while maintaining territorial integrity, projecting power, and producing high cultural output}. Such civilizations would prove that demographic/economic/social health are not necessary conditions for sustained geopolitical viability.
\item \textbf{Current status:} No clear counterexamples identified. Late-stage empires (Late Rome, Late Qing) exhibited low V but rapidly collapsed or were conquered within 1-2 generations of reaching low-V state.
\end{itemize}

\textbf{3. The C-Axis Falsification Protocol (Constraint):}
\begin{itemize}
\item \textbf{The Hypothesis:} A polity's ability to act is meaningfully constrained by its position within the geopolitical power structure. Vassals have less agency than hegemons.
\item \textbf{The Falsification Condition:} The discovery of \textbf{systematic patterns where measured C-axis values (hegemon vs vassal status) show no correlation with observed policy autonomy or strategic choices}. For example: if vassal states (C \textless{} -0.5) consistently demonstrate the same range of strategic options as hegemons (C \textgreater{} +0.5), the C-axis construct lacks predictive validity.
\item \textbf{Measurement:} Compare policy independence indices (trade agreements, military alliances, domestic legislation) across civilizations with varying C-axis scores. If correlation between C and autonomy is weak (r \textless{} 0.3), C-axis adds no explanatory power.
\end{itemize}

\subsection{\texorpdfstring{\textbf{III. Falsifying the Deep Physics (The Bedrock)}}{III. Falsifying the Deep Physics (The Bedrock)}}\label{iii.-falsifying-the-deep-physics-the-bedrock}

These are the most fundamental claims. To falsify one of these is to shatter the entire edifice.

\textbf{1. The Holographic Hypothesis Falsification Protocols}

The holographic hypothesis claims that axiological patterns at the civilizational scale are isomorphic with patterns at the neurological scale and potentially universal across intelligent systems. This hypothesis can be challenged on multiple fronts:

\vspace{0.5em}

\textbf{CRITICAL DEPENDENCY:} The holographic principle rests on the validity of hemispheric specialization (McGilchrist's model of functional duality in brain architecture). This is a load-bearing assumption with medium epistemic confidence (Tier 2). If neuroscientific consensus decisively overturns the hemispheric specialization model, the holographic argument requires fundamental revision—the civilizational patterns would remain observable, but their claimed universality (grounding in computational necessity rather than cultural contingency) would be weakened. The framework's diagnostic utility (Parts I-II, IV) remains intact regardless; the deep physics claims (Part III, holographic principle) depend on this neuroscientific foundation.

\vspace{0.5em}

\textbf{1a. The Neurological Test (Test of Isomorphism):}
\begin{itemize}
\item \textbf{The Claim:} The same fundamental, dialectical patterns (The Trinity of Tensions) repeat at the level of civilization, neuroscience, and computation.
\item \textbf{The Falsification:} The hypothesis would be invalidated if \textbf{future neuroscientific research decisively overturns the McGilchrist model of hemispheric specialization}. If it is proven that the brain does not operate on a fundamental, functional duality, then our claim of isomorphism collapses, and the apparent pattern is revealed to be a mere coincidence or cultural artifact rather than universal computational necessity.
\end{itemize}

\textbf{1b. The Cross-Species Test (The Alien Test):}
\begin{itemize}
\item \textbf{The Claim:} The axiological patterns of civilization (The Grand Cycle) are isomorphic with the functional architecture of the human brain (The Hemispheric Dialectic) because both reflect universal computational constraints faced by any intelligent system.
\item \textbf{The Falsification:} The discovery of \textbf{≥30\% of successful, durable civilizations (Ω \textgreater{} 0.5, sustained \textgreater{}5 generations) whose axiological signatures require fundamentally non-bipolar dimensional structures}. Specifically: if analysis of \textgreater{}20 diverse historical civilizations reveals that \textgreater{}6 require ``tripolar'' or ``quadripolar'' axes (not reducible to SORT), then our claim that bipolar dialectic is universal is falsified. This would prove the model is a local artifact of human neurobiology, not universal computational necessity.
\end{itemize}

\textbf{2. The Trinity of Tensions Falsification Condition (The Fourth Tension Test)}
\begin{itemize}
\item \textbf{The Claim:} There are three and only three fundamental, orthogonal problems of existence (World, Time, Self).
\item \textbf{The Falsification:} The hypothesis would be invalidated if a thinker could propose a \textbf{Fourth, Irreducible Tension} that is as fundamental as the first three and cannot be explained as a sub-problem or a combination of them. This would prove that our foundational map of the problem space is incomplete.
\end{itemize}

\textbf{3. The Environmental Selection Falsification Condition (The Test of the Prime Mover)}
\begin{itemize}
\item \textbf{The Claim:} The Grand Cycle is driven by the impersonal selection pressures of Scarcity and Abundance.
\item \textbf{The Falsification:} The hypothesis would be invalidated if \textbf{≥5 civilizations across diverse cultural contexts and time periods} exhibit inverted environmental responses:
\item Polities in profound, long-term \textbf{Abundance} (GDP/capita \textgreater{}2× subsistence, \textgreater{}3 generations of peace) spontaneously generate \textbf{hard, martial, Foundry axiology (T+ \textgreater{} +0.5, S+ \textgreater{} +0.5)}
\item Polities in brutal, existential \textbf{Scarcity} (famine, war, resource collapse) spontaneously generate \textbf{soft, tolerant, Hospice axiology (T- \textless{} -0.3, therapeutic culture)}
\end{itemize}
This would disprove that environmental selection is the prime axiological selector. Isolated anomalies (1-2 cases) do not falsify; systematic inverted patterns (≥5 cases across diverse contexts) would.

\textbf{4. The Three-Layer Architecture Falsification Protocols (CH15)}

CH15 claims that durable, complex polities require exactly three differentiated functional layers (Heart/Skeleton/Head) to solve the Trinity of Problems (Continuity/Constraint/Direction), and that these layers must maintain specific axiological alignments. The following protocols test both the architectural necessity and the alignment requirements:

\textbf{4a. The Architectural Necessity Test (The Flat Polity):}
\begin{itemize}
\item \textbf{The Hypothesis:} Durable, complex polities require exactly three differentiated functional layers. Systems with fewer layers fail through brittleness or incoherence; systems with more layers introduce parasitic complexity without solving new fundamental problems.
\item \textbf{The Falsification Condition (The Alternative Architecture):} The discovery of \textbf{durable, high-performing civilizations (sustained Ω > 0.5, V > 6.0 across multiple generations) with clearly non-three-layer architecture}:
\begin{itemize}
\item A one-layer system (undifferentiated monolith) sustaining long-term viability
\item A two-layer system successfully solving all three Trinity problems without the third layer
\item A four-or-more-layer system where the additional layer(s) solve genuinely distinct fundamental problems not reducible to Heart/Skeleton/Head functions
\end{itemize}
This would falsify the claim that three-layer architecture is a universal structural necessity, proving it is instead a contingent historical pattern or optimal-but-not-required design.
\end{itemize}

\textbf{4b. The T-Axis Alignment Test (The Revolutionary Substrate):}
\begin{itemize}
\item \textbf{The Hypothesis:} Stable Foundries require T+ (Metamorphic) Head with T- (Homeostatic) Heart, mediated by T- Skeleton. A T+ Heart creates unsustainable revolutionary fervor in the substrate; a T- Head produces civilizational stagnation.
\item \textbf{The Falsification Condition:} Discovery of \textbf{durable, high-performing Foundries (Ω > 0.5, Α+ > 0.5, sustained across multiple generations) with T+ substrate populations (Heart) demanding constant revolutionary transformation}. If such civilizations maintain stability despite bottom-up revolutionary pressure, the claim that Hearts must be Homeostatic is falsified.
\end{itemize}

\textbf{4c. The R-Axis Alignment Test (The Mythos Sovereign):}
\begin{itemize}
\item \textbf{The Hypothesis:} Stable Foundries require R+ (Gnostic) Head with R- (Mythos) Heart. The Head must reality-test while the Heart provides meaning and coherence. Inverting this (R- Head, R+ Heart) produces either theocratic brittleness or meaningless technocracy.
\item \textbf{The Falsification Condition:} Discovery of \textbf{durable Foundries (sustained high performance across multiple generations) governed by explicitly R- leadership (theocratic, tradition-bound, anti-empirical) ruling over R+ populations demanding Gnostic governance}. If such inverted configurations prove stable and high-performing, the R-axis alignment claim is falsified.
\end{itemize}

\textbf{4d. The S-Axis Alignment Test (The Individualist Sovereign):}
\begin{itemize}
\item \textbf{The Hypothesis:} Stable Foundries require S+ (Collectivist) Head serving the polity's interests with S- (Individualist) Heart preserving personal agency. The Head must serve "We"; the Heart must protect "I." Inverting this produces either tyranny (S+ substrate) or anarchic dysfunction (S- Head).
\item \textbf{The Falsification Condition:} Discovery of \textbf{durable Foundries with S- leadership (individualist, self-serving sovereigns) governing S+ populations (collectivist substrate) that nonetheless maintain high Ω and Α+ across multiple generations}. If purely self-interested leadership can sustainably govern collectivist populations, the S-axis alignment claim is falsified.
\end{itemize}

These architectural and alignment claims are among the most specific and novel engineering principles the framework offers. Finding \textbf{≥3 stable civilizations (Ω \textgreater{} 0.5, V \textgreater{} 6.0, sustained \textgreater{}5 generations)} with alternative architectures (non-three-layer) or inverted layer alignments would require fundamental revision of the Polytheistic Governance principle.

\subsection{\texorpdfstring{\textbf{IV. Falsifying the V1.0 Diagnostic Framework: Ω-Α Dynamics}}{IV. Falsifying the V1.0 Diagnostic Framework: Ω-Α Dynamics}}\label{iv.-falsifying-v1.0-framework}

The V1.0 framework used throughout this book operates on directly observable Ω-Α dynamics (State Coherence and Action Vector). The following protocols falsify the core V1.0 diagnostic model. \textbf{Note:} The V0.1 State Potentials (Ψ/Ν/Κ formulas) are research explorations with their own falsification protocols documented in \Cref{app:research}.

\textbf{1. The Iron Law of Coherence (The Forbidden Quadrant Test):}
\begin{itemize}
\item \textbf{The Hypothesis:} The Top-Left Quadrant of the Α-Ω Phase Space (Low Ω, High Α+) is a Forbidden Zone. A polity cannot sustain high Syntropic Action (order-creation) while in a state of low Coherence (internal conflict).
\item \textbf{The Falsification Condition (The Creative Cauldron):} Find a single clear historical example of a deeply incoherent polity (civil war, tribal fragmentation, Ω < 0.4) that simultaneously executed large-scale Syntropic projects (built empires, created lasting infrastructure, Α+ > 0.5) for multiple decades.
\item \textbf{Current Status:} No clear counterexample found in systematic historical survey.
\end{itemize}

\subsection{\texorpdfstring{\textbf{V. Falsifying the Grand Narratives: The Deep Theories}}{V. Falsifying the Grand Narratives: The Deep Theories}}\label{v.-falsifying-the-grand-narratives-the-deep-theories}

Finally, stating the conditions that would falsify the broadest, most speculative, but most important syntheses.

\textbf{1. The Grand Cycle Falsification Protocol:}
\begin{itemize}
\item \textbf{The Hypothesis:} History is driven by a cycle in which Scarcity selects for the Foundry (ALPHA), which creates Abundance, which selects for the Hospice (BETA), which creates a new Scarcity.
\item \textbf{The Falsification Condition (The Stable Utopia):} The discovery of a \textbf{civilization that successfully achieves a State of Abundance and \emph{remains} there for >500 years (>20 generations) without succumbing to axiological decay}. Specifically: maintaining Ω > 0.5, Α+ > 0.3, and T+ > +0.2 continuously while in economic abundance. A civilization that can be simultaneously rich, safe, and permanently Metamorphic. Such a discovery would disprove our entire theory of the Four Horsemen and prove that a stable, pro-Aliveness utopia is possible. It would be the most joyful falsification imaginable.
\end{itemize}

\textbf{2. The Axiological Entanglement Falsification Protocol:}
\begin{itemize}
\item \textbf{The Hypothesis:} Large-scale, expansive empire-building (high T+, high Α+) requires collective mobilization (S+). Pure Individualism (S-) cannot generate the coordinated sacrifice and group cohesion necessary for sustained imperial conquest and administration.
\item \textbf{The Falsification Condition (The Individualist Empire):} The discovery of \textbf{large-scale, expansive, empire-building civilizations that successfully operated across multiple generations on a foundation of pure Individualism (S- ≈ -0.7 or lower)}, without developing strong collective identity or demanding collective sacrifice. Civilizations of pure mercenaries and rational self-interest that nonetheless conquered territories, administered provinces, and projected power across continents. Such discoveries would disprove our claim that S and T axes are axiologically entangled at the extremes.
\end{itemize}

\textbf{3. The Foundry Imperative Falsification Protocols:}

\textit{(Note: These protocols test the historical consequences and typological claims derived from the 3-Layer Architecture. For protocols that test the necessity of the 3-Layer Architecture itself, see \Cref{iii.-falsifying-the-deep-physics-the-bedrock}, Protocol 4.)}

CH13 presents an exhaustive survey claiming only ALPHA (Foundry/T+) states are durably viable, structured as six fundamental configurations. The following protocols test each major claim:

\textbf{3a. The Minimum T+ Threshold (Swiss Floor):}
\begin{itemize}
\item \textbf{The Hypothesis:} There exists a minimum viable T+ threshold (approximately +0.2, demonstrated by Switzerland) below which states inevitably enter Hospice decay patterns via the Four Horsemen mechanism. Pure T- (Homeostatic) polities cannot sustain long-term stability.
\item \textbf{The Falsification Condition (The Stable Hospice):} The discovery of \textbf{civilizations that maintained T-axis values below +0.2 (or demonstrably negative T-axis values) while sustaining high Coherence (Ω > 0.5) and positive Action (Α+ > 0.3) for >10 generations (>250 years) without external conquest}. If multiple such examples exist with consistent patterns, this would falsify the claim that T+ is categorically required for durable stability and that the Swiss configuration represents a hard floor rather than one point on a continuum.
\end{itemize}

\textbf{3b. The Gnostic Imperative (Universal R+ Requirement):}
\begin{itemize}
\item \textbf{The Hypothesis:} All durable Foundries require predominantly R+ (Gnostic) orientation because T+ (Metamorphosis) generates complex, multi-generational projects that brutally test reality-contact. R- (Mythos-driven) variants are ``Brittle Foundries''---capable of short bursts but systematically shattered by more Gnostically competent powers.
\item \textbf{The Falsification Condition (The Successful Mythos Empire):} The discovery of \textbf{civilizations with measured R-axis values substantially below zero (R < -0.3) that nonetheless sustained high performance (Ω > 0.5, Α+ > 0.5) for multiple centuries (>5 generations) without being conquered by R+ powers}. This would prove that Mythos-driven orientation can produce durable complex civilizations, falsifying the claim that Gnosis is a near-universal requirement for Foundry viability.
\end{itemize}

\textbf{3c. The Defensive S/O Synergy Constraint:}
\begin{itemize}
\item \textbf{The Hypothesis:} For Defensive Foundries, viable S/O pairings are constrained by physical synergy requirements: O- enables S- (Citadel), O+ requires S+ (Fortress), O≈0 produces S≈0 (Confederal). Mismatched configurations [O+/S-] or [O-/S+] are internally contradictory and dynamically unstable.
\item \textbf{The Falsification Condition (The Contradictory Pairing):} The discovery of \textbf{stable Defensive Foundries (sustained viability across multiple generations) with contradictory S/O pairings}:
\begin{itemize}
\item O+ (centralized Design) paired with S- ≈ -0.5 or below (strong Individualism)
\item O- (decentralized Emergence) paired with S+ ≈ +0.5 or above (strong Collectivism)
\end{itemize}
Multiple examples of such configurations maintaining stability would falsify the claim that S/O synergy is a binding physical constraint rather than a statistical tendency.
\end{itemize}

\textbf{3d. The Six-Type Completeness (Typological Exhaustiveness):}
\begin{itemize}
\item \textbf{The Hypothesis:} The viable Foundry state space consists of exactly six fundamental configurations, determined by Telos Scope (Expansive vs. Defensive) × O-Axis position (O-, O≈0, O+). All stable, high-performing civilizations fit one of these six archetypes within specified SORT ranges. (See CH13, Table at line 508-523 for complete specifications.)
\item \textbf{The Falsification Condition (The Seventh Archetype):} The discovery of \textbf{systematic patterns of stable civilizations (Ω > 0.5, Α+ > 0.3 sustained across generations) that fall significantly outside the characteristic SORT ranges of all six types}. Examples that would falsify:
\begin{itemize}
\item A seventh distinct Telos Scope beyond Expansive/Defensive (e.g., ``Oscillating'' states that regularly alternate directions)
\item A fourth viable O-axis position beyond O-/O≈0/O+ (e.g., ``Dynamic'' states that cycle between positions)
\item Stable configurations at SORT coordinates forbidden by the typology (e.g., Defensive/O+ with S- < -0.5)
\end{itemize}
Isolated outliers (1-2 cases) or short-lived anomalies (\textless{}3 generations) do not falsify; \textbf{≥4 stable civilizations} exhibiting viability outside the six-type space for \textgreater{}5 generations would require fundamental revision or expansion of the typology.
\end{itemize}

\subsection{\texorpdfstring{\textbf{VI. The Research Invitation}}{VI. The Research Invitation}}\label{vi.-the-research-invitation}

This framework emerged through AI-mediated synthesis—pattern recognition across compressed historical knowledge. Such synthesis excels at hypothesis generation but is not a substitute for empirical validation.

\textbf{Division of labor:} I am a synthesizer, not a validator. I have generated frameworks and testable hypotheses. I have not performed systematic data collection, statistical analysis, or quantitative measurement. Both roles are necessary. Synthesis without validation is speculation.

\textbf{Success metrics:} Peer-reviewed papers (positive or negative), dissertations, real-world implementations, competing frameworks, SORT adoption, or major claims falsified. Better comprehensive synthesis partially wrong than safe incrementalism precisely irrelevant.

\subsection{\texorpdfstring{\textbf{VII. Proposed Falsification Roadmap for Future Personal-Scale Research (\Cref{part:integrated-human})}}{VII. Proposed Falsification Roadmap for Future Personal-Scale Research (Part V)}}\label{vii.-falsifying-part-v}

The framework claims scale-invariance: the same physics governs civilizations (\cref{part:diagnosis,part:autopsy,part:source-code,part:blueprint}) and individuals (\Cref{part:integrated-human}). For this claim to be credible, \Cref{part:integrated-human} must be falsifiable in principle using protocols structurally similar to those validating civilizational SORT.

\textbf{Current Epistemic Status:} As of publication, Part V is \textbf{unfalsifiable in practice} due to the absence of a validated pSORT assessment instrument. The protocols below represent a research roadmap for future empirical validation, demonstrating that Part V \textit{could} be tested if the necessary measurement tools are developed. Until then, Part V should be evaluated on theoretical coherence, consilience with existing research, and pragmatic N-of-1 utility—not empirical validation.

\textbf{1. The pSORT Dimensionality Test}

\textbf{Claim:} Personal axiology captured by four SORT dimensions. \textbf{Falsification:} Large-scale psychometric studies (N>1000) consistently reveal irreconcilable dimensional structure (e.g., 5/7/12 factors with no SORT mapping). \textbf{Protocol:} Confirmatory factor analysis on validated pSORT instrument vs. Big Five/HEXACO. Threshold: Four-factor model shows RMSEA > 0.08, CFI < 0.90.

\textbf{2. The Mask-Coherence Hypothesis Test}
\begin{itemize}
\item \textbf{The Claim:} The "Mask" (sustained divergence between native pSORT and performed pSORT) causes measurable reduction in personal coherence ($\Omega_p$), manifesting as anxiety, burnout, and reduced life satisfaction.
\item \textbf{The Falsification:} The Mask mechanism would be falsified if individuals with large self-reported native vs. performed pSORT divergence show NO statistically significant difference in coherence-related outcomes compared to individuals with minimal divergence.
\item \textbf{Test Protocol:} Measure native vs. performed pSORT divergence via validated assessment. Measure coherence outcomes via established instruments: Maslach Burnout Inventory (emotional exhaustion subscale), Generalized Anxiety Disorder scale (GAD-7), Satisfaction with Life Scale (SWLS). Control for work hours, socioeconomic status, major life events. If high-divergence individuals show no meaningful difference in coherence measures from low-divergence individuals, Mask-as-pathology hypothesis is falsified. (Specific divergence thresholds and minimum effect sizes require empirical calibration once validated instruments exist.)
\end{itemize}

\textbf{3. The Integration Intervention Test}
\begin{itemize}
\item \textbf{The Claim:} Individuals who consciously adopt the Internal Polity architecture (3-layer personal governance: Heart/Skeleton/Head as described in \Cref{ch:integrated-human}) will show measurable increases in personal coherence ($\Omega_p$) and constructive action ($\Alpha_p$+) compared to baseline.
\item \textbf{The Falsification:} The prescriptive utility of the Internal Polity framework would be falsified if controlled intervention studies show NO significant improvement in intervention group compared to control group.
\item \textbf{Test Protocol:} Randomized controlled trial. Intervention group: structured program implementing Internal Polity architecture (regular practice of Heart/Skeleton/Head differentiation, Great Work identification, Circuit-Breaker protocols). Control group: matched on baseline $\Omega_p$ and demographics, receives generic self-improvement content. Measure $\Omega_p$ (via pSORT variance + life satisfaction) and $\Alpha_p$ (via self-reported meaningful output + independent evaluations of creative/professional work). If intervention group shows no meaningful improvement over control group on combined coherence and action metrics, prescriptive framework is falsified. (Specific program duration and minimum effect sizes require empirical calibration.)
\end{itemize}

\textbf{4. The Implicit Treaty Collision Test}
\begin{itemize}
\item \textbf{The Claim:} Romantic partnerships with high pSORT divergence experience predictably higher conflict rates than convergent partnerships, due to incompatible "Implicit Treaties" (unconscious axiological constitutions).
\item \textbf{The Falsification:} The Implicit Treaty collision model would be falsified if measured pSORT divergence shows NO correlation with relationship conflict patterns.
\item \textbf{Test Protocol:} Measure couple pSORT divergence via validated assessment. Measure conflict via Gottman Conflict Scale (weekly conflict frequency + severity). Control for relationship duration, age, children, financial stress. If high-divergence couples show no meaningful difference in conflict rates from low-divergence couples, Implicit Treaty model is falsified. (Specific divergence thresholds and expected effect sizes require empirical calibration once validated instruments exist.)
\end{itemize}

\textbf{5. The Neurodivergence-Mask Correlation Test}
\begin{itemize}
\item \textbf{The Claim:} "Systemizers" (high R+/O+ natives, often autistic/ADHD) in "Empathizer"-dominant cultures (high R-/S+ institutional norms) show higher Mask prevalence and lower $\Omega_p$ than Systemizers in Systemizer-tolerant cultures, due to forced axiological suppression.
\item \textbf{The Falsification:} The environmental mismatch model would be falsified if cross-cultural comparisons show NO correlation between cultural pSORT demands and Mask prevalence/coherence outcomes in neurodivergent populations.
\item \textbf{Test Protocol:} Cross-cultural study comparing Systemizer Quotient (Baron-Cohen) scores, Mask prevalence (via Energy Test + Childhood Test convergence), and anxiety/burnout measures across cultures with different dominant axiologies. Control for baseline neurodivergence rates, socioeconomic development. If Mask rates and coherence outcomes show no correlation with cultural pSORT mismatch, environmental model is falsified. (Specific correlation thresholds and minimum effect sizes require empirical calibration.)
\end{itemize}

\textbf{6. The Unmasking Recovery Test}
\begin{itemize}
\item \textbf{The Claim:} Individuals who successfully complete the Unmasking protocol (\Cref{ch:integrated-human}: identifying and releasing counterfeit pSORT signature) show measurable recovery in $\Omega_p$ and reduction in burnout/anxiety symptoms.
\item \textbf{The Falsification:} The Unmasking protocol would be falsified as therapeutic intervention if individuals who complete it show NO improvement in coherence measures compared to their pre-Unmasking baseline or compared to matched controls.
\item \textbf{Test Protocol:} Longitudinal study tracking individuals through Unmasking process. Measure baseline $\Omega_p$, native vs. performed pSORT divergence, burnout (MBI), anxiety (GAD-7). Re-measure post-Unmasking. Compare to waitlist control group. If Unmasking group shows no meaningful improvement in coherence measures compared to controls, protocol lacks therapeutic validity. (Specific follow-up timing and minimum effect sizes require empirical calibration.)
\end{itemize}

\textbf{The Validation Gap:}

All six protocols above require a \textbf{validated pSORT assessment instrument}, which does not yet exist. This is the fundamental reason Part V is currently unfalsifiable in practice. Current Part V claims rest on:
\begin{itemize}
\item Qualitative N-of-1 heuristics (Energy Test, Childhood Test, Crisis Test)
\item Consilience with existing research (attachment theory, neurodivergent self-reports, Gottman's marital research)
\item Internal consistency with \cref{part:diagnosis,part:autopsy,part:source-code,part:blueprint}
\item Pragmatic utility in individual experimentation
\end{itemize}

Developing and validating a pSORT assessment instrument is the critical prerequisite for moving Part V from Tier 2-3 theoretical framework to Tier 1 empirically testable science. Until validated instruments exist, Part V cannot be subjected to rigorous falsification tests and should be evaluated primarily on theoretical coherence, consilience, and pragmatic N-of-1 utility—not statistical validation.

\textbf{The Scale-Invariance Test:}

The ultimate test of the holographic claim is whether the SAME falsification principles apply at both scales:
\begin{itemize}
\item If civilizational SORT is falsifiable but personal pSORT is not → holographic claim fails (personal scale is unfalsifiable metaphor)
\item If both are falsifiable but use incompatible methodologies → holographic claim is weakened (suggests different physics at different scales)
\item If both are falsifiable via structurally similar protocols → holographic claim is strengthened (suggests genuine scale-invariance)
\end{itemize}

The protocols above demonstrate that \Cref{part:integrated-human} IS falsifiable using the same conceptual structure as \cref{part:diagnosis,part:autopsy,part:source-code,part:blueprint}: dimensionality tests, correlation predictions, intervention outcomes, cross-cultural patterns. The physics may be universal even if the measurement tools are still being built.

---

This is the complete falsification architecture and research agenda. We have defined both what would break the framework at BOTH scales (civilizational and personal) AND the positive work needed to validate, refine, and extend it.
