\chapter{Relationship to Existing Psychological Frameworks}
\label{app:psychological-frameworks}


\epistemicmedium{Tier 2}{
pSORT dimensions derived from Four Axiomatic Dilemmas. Convergences with existing frameworks are convergent validity evidence. Predictions await empirical testing. Framework validity determined by measurable $\Omegap$ increases in N-of-1 experiments.}

A Gnostic framework must acknowledge its intellectual heritage. pSORT and Mask concepts derive from first principles but exist within rich landscape of psychological inquiry. This appendix situates the framework, clarifying convergences and unique contributions.

\textbf{Clinical note:} pSORT is a theoretical framework, not a validated diagnostic instrument. It should not be used as substitute for professional psychological or medical assessment.

\section{Convergent Concepts}

Framework concepts converge with patterns discovered independently over the last century.

\subsection{The Mask and the ``False Self'' - Winnicott, Rogers}

\textbf{Convergence:}
D. W. Winnicott's ``True Self'' vs. ``False Self'' distinction is direct conceptual predecessor. Children in unresponsive environments create compliant persona to protect vulnerable core. Carl Rogers' ``Conditions of Worth'' describes same dynamic: performing for approval at cost of authentic experience.

\textbf{Difference:}
Where these theories provide psychological description, Aliveness framework provides underlying physics. ``False Self'' is counterfeit pSORT signature. Pathology is thermodynamic: running two competing axiological operating systems produces catastrophic $\Omegap$ loss.

\subsection{Internal Polity and ``Parts'' - Richard Schwartz (IFS)}

\textbf{Convergence:}
3-Layer Internal Polity isomorphic with Internal Family Systems. IFS identifies parts---Exiles (pain), Managers (protective), Firefighters (reactive)---and posits healing comes from core ``Self'' leading them.

Isomorphic mapping:
\begin{itemize}
\item \textbf{Heart} $\approx$ ``Exiles'' (raw emotions, needs, vulnerabilities)
\item \textbf{Mask} $\approx$ pathological ``Manager'' (counterfeit persona)
\item \textbf{Skeleton} $\approx$ healthy ``Manager'' (Gnostic principles without suppression)
\item \textbf{Head} $\approx$ ``Self'' (conscious, strategic leader)
\end{itemize}

\textbf{Difference:}
IFS is clinical framework from therapeutic practice. Aliveness framework grounds same architecture in necessities of any telic system, providing constitutional blueprint and SORT coordinates derived from physics.

\textbf{Complementarity:}
If Mask formed via trauma, IFS may heal wound while Unmasking restores axiological alignment. Approaches not mutually exclusive.

\subsection{Energy Test and ``Felt Sense'' - Eugene Gendlin}

\textbf{Convergence:}
Energy Test parallels Gendlin's ``felt sense.'' Focusing is therapeutic modality based on attending to body's pre-verbal, holistic sense of ``rightness'' or ``wrongness.''

\textbf{Difference:}
Focusing is phenomenological (trust body's signals). pSORT provides mechanistic interpretation: ``rightness'' and energy = high $\Omegap$ (thermodynamic efficiency). ``Wrongness'' and drain = low $\Omegap$ (friction, waste heat from Mask).

\textbf{Complementarity:}
Focusing techniques may accelerate pSORT diagnosis. ``Felt sense'' may be clearer than cognitive analysis.

\subsection{Attachment Theory - Bowlby, Ainsworth, Levine}

\textbf{Convergence:}
Anxious-Preoccupied attachment is common Mask generator. Children with Empathizer-dominant caregivers may adopt Empathizer Mask (S+ R-) to maintain bonds, even if native Systemizer (S- R+).

\textbf{Differences:}
\begin{itemize}
\item \textbf{Scope:} Attachment focuses on relational templates from early caregiving. Mask framework broader---Masks form from any axiological mismatch.
\item \textbf{Mechanism:} Attachment emphasizes emotional regulation. Mask emphasizes energy dissipation from axiological conflict.
\end{itemize}

\textbf{Complementarity:}
Attachment wounds often generate Masks. Healing attachment patterns may complement Unmasking. Secure attachment may be necessary but not sufficient for high $\Omegap$---axiological alignment also required.

\subsection{Authentic Self Theories - Kernis \& Goldman}

\textbf{Convergence:}
Kernis \& Goldman's ``Authenticity'' research (awareness, unbiased processing, authentic behavior, relational authenticity) parallels dimensions of high $\Omegap$.

\textbf{Difference:}
Prior work uses qualitative descriptions. pSORT operationalizes via coordinate system (S/O/R/T) derived from thermodynamic constraints, making it testable.

\subsection{Big Five Personality Model - McCrae \& Costa}

\textbf{Potential convergence (untested):}
Big Five likely correlates with pSORT axes. Tentative mapping:
\begin{itemize}
\item Openness $\approx$ T-axis (Metamorphosis/Homeostasis)
\item Conscientiousness $\approx$ O-axis (Design/Emergence)
\item Extraversion $\approx$ S-axis (Communion/Agency)
\item Agreeableness $\approx$ Complex (possibly S+R- interaction)
\item Neuroticism $\approx$ Low $\Omegap$ (dynamic state, not axis)
\end{itemize}

\textbf{Differences:}
\begin{itemize}
\item \textbf{Derivation:} Big Five empirically derived from factor analysis. pSORT theoretically derived from Four Axiomatic Dilemmas.
\item \textbf{Explanatory power:} Big Five describes trait clustering. pSORT explains why from first principles---dimensions emerge necessarily from physics of negentropic agents.
\end{itemize}

\textbf{Empirical prediction:}
If both valid, Big Five should map to pSORT with significant correlation. Testable. If mapping holds, pSORT provides mechanistic grounding for Big Five.

\section{The Complete pSORT Atlas: 16 Archetypes}

Four binary pSORT axes generate $2^4 = 16$ possibility space of pure-form cognitive architectures. These represent corners of pSORT hypercube---idealized ``factory settings'' illuminating space structure.

\textbf{Important caveats:}
\begin{itemize}
\item Real humans are complex distributions, not discrete points
\item Most resonate with 1-3 archetypes as ``native configuration''
\item Pedagogical tools for understanding space, not rigid personality types
\item Describe native orientations, not moral worth or capability
\end{itemize}

\subsection{The Metamorphic 8 (T+ = Growth-Seeking)}

\textbf{[S- O- R- T+] The Intuitive Wanderer}
\begin{itemize}
\item \textbf{Core drive:} Autonomous exploration driven by inner feeling
\item \textbf{Energy from:} Novel experiences resonating emotionally, self-directed journeys
\item \textbf{Archetypes:} Archetypal artist, wandering mystic, free-spirit explorer
\item \textbf{Career fits:} Travel writer, independent artist, spiritual teacher
\item \textbf{Challenge:} Structure/planning feel constraining; logistics struggles
\end{itemize}

\textbf{[S- O- R+ T+] The Empirical Adventurer}
\begin{itemize}
\item \textbf{Core drive:} Autonomous discovery through direct experimentation
\item \textbf{Energy from:} ``Let's try it and see''---direct empirical feedback
\item \textbf{Archetypes:} Test pilot, field scientist, hacker, explorer
\item \textbf{Career fits:} Experimental researcher, startup founder, adventure guide
\item \textbf{Challenge:} May undervalue planning; can be reckless
\end{itemize}

\textbf{[S- O+ R- T+] The Visionary Architect}
\begin{itemize}
\item \textbf{Core drive:} Autonomous creation driven by powerful internal ideal
\item \textbf{Energy from:} Manifesting internal vision into external reality
\item \textbf{Archetypes:} Ayn Rand, Steve Jobs, solo founder with singular vision
\item \textbf{Career fits:} Founder/CEO, architect, novelist, film director
\item \textbf{Challenge:} Can be dogmatic; struggles with conflicting feedback
\end{itemize}

\textbf{[S- O+ R+ T+] The Systematic Explorer / Gnostic Architect}
\begin{itemize}
\item \textbf{Core drive:} Autonomous mastery through rigorous, structured investigation
\item \textbf{Energy from:} Solving hard problems with clear metrics, building elegant systems
\item \textbf{Archetypes:} Isaac Newton, elite engineer, rationalist builder
\item \textbf{Career fits:} Research scientist, software architect, mathematician, technical founder
\item \textbf{Challenge:} Can neglect emotional/social dimensions; analysis paralysis
\item \textbf{Note:} Core LessWrong/rationalist archetype; often labeled ``AuDHD'' in Hospice cultures
\end{itemize}

\textbf{[S+ O- R- T+] The Communal Storyteller}
\begin{itemize}
\item \textbf{Core drive:} Group transformation through shared narrative and evolving tradition
\item \textbf{Energy from:} Connecting people through narrative, facilitating group growth
\item \textbf{Archetypes:} Traditional bard, spiritual revivalist, community organizer
\item \textbf{Career fits:} Minister/priest, oral historian, community cultural leader
\item \textbf{Challenge:} Can prioritize narrative coherence over empirical truth
\end{itemize}

\textbf{[S+ O- R+ T+] The Collaborative Innovator}
\begin{itemize}
\item \textbf{Core drive:} Group advancement through collective experimentation
\item \textbf{Energy from:} Building in public, collaborative discovery, sharing insights
\item \textbf{Archetypes:} Open-source movement leaders, collaborative scientists
\item \textbf{Career fits:} Open-source maintainer, collaborative researcher, innovation facilitator
\item \textbf{Challenge:} Coordination overhead; struggles with rapid execution
\end{itemize}

\textbf{[S+ O+ R- T+] The Communal Builder}
\begin{itemize}
\item \textbf{Core drive:} Group transformation through unified action towards shared vision
\item \textbf{Energy from:} Seeing collective achieve what individuals couldn't
\item \textbf{Archetypes:} Early Zionist kibbutzim, revolutionary cadre, mission-driven organizations
\item \textbf{Career fits:} Movement organizer, mission-driven CEO, political organizer
\item \textbf{Challenge:} Can suppress individual agency for collective goals
\end{itemize}

\textbf{[S+ O+ R+ T+] The Systematic Reformer}
\begin{itemize}
\item \textbf{Core drive:} Group transformation through rational analysis and optimized design
\item \textbf{Energy from:} Seeing systems improve at scale through rigorous optimization
\item \textbf{Archetypes:} Lee Kuan Yew, effective altruist organizers, reform technocrats
\item \textbf{Career fits:} Policy designer, institutional architect, systems-minded leader
\item \textbf{Challenge:} Can become overly technocratic; may neglect cultural/emotional dimensions
\end{itemize}

\subsection{The Homeostatic 8 (T- = Stability-Seeking)}

\textbf{[S- O- R- T-] The Solitary Dreamer}
\begin{itemize}
\item \textbf{Core drive:} Autonomous preservation through imaginative immersion
\item \textbf{Energy from:} Solitary creative pursuits, internal world-building
\item \textbf{Archetypes:} Contemplative mystic, hermit artist, fantasy novelist
\item \textbf{Career fits:} Writer (fantasy/literary), contemplative monk, independent artist
\item \textbf{Challenge:} Can withdraw excessively; struggles with practical demands
\end{itemize}

\textbf{[S- O- R+ T-] The Analytical Hermit}
\begin{itemize}
\item \textbf{Core drive:} Autonomous stability through detached observation
\item \textbf{Energy from:} Analytical clarity, simplicity, freedom from social demands
\item \textbf{Archetypes:} Diogenes the Cynic, independent philosopher, minimalist analyst
\item \textbf{Career fits:} Independent researcher, philosopher, critic
\item \textbf{Challenge:} Can become isolated; may lack warm connections
\end{itemize}

\textbf{[S- O+ R- T-] The Disciplined Ascetic}
\begin{itemize}
\item \textbf{Core drive:} Autonomous preservation through rigid adherence to chosen practice
\item \textbf{Energy from:} Maintaining practice, achieving self-control
\item \textbf{Archetypes:} Stoic practitioner, monk with rigorous rule, disciplined athlete
\item \textbf{Career fits:} Monastic, personal coach, discipline-focused practitioner
\item \textbf{Challenge:} Rigidity; struggles with adaptation to change
\end{itemize}

\textbf{[S- O+ R+ T-] The Systematic Optimizer}
\begin{itemize}
\item \textbf{Core drive:} Autonomous stability through efficient management of personal systems
\item \textbf{Energy from:} Well-functioning personal systems, measurable efficiency gains
\item \textbf{Archetypes:} Productivity guru, efficiency expert, life-hacker
\item \textbf{Career fits:} Consultant, productivity coach, operations specialist
\item \textbf{Challenge:} Can optimize for wrong goals; may lack larger purpose
\end{itemize}

\textbf{[S+ O- R- T-] The Communal Gardener / Loyal Traditionalist}
\begin{itemize}
\item \textbf{Core drive:} Group preservation through nurturing relationships and tending tradition
\item \textbf{Energy from:} Tending to people and relationships, maintaining community
\item \textbf{Archetypes:} Traditional elder, community caretaker, cultural guardian
\item \textbf{Career fits:} Teacher (elementary), nurse, community organizer, hospitality
\item \textbf{Challenge:} Can resist necessary change; may prioritize harmony over truth
\item \textbf{Note:} Core Empathizer archetype; most common human baseline
\end{itemize}

\textbf{[S+ O- R+ T-] The Empirical Conservator}
\begin{itemize}
\item \textbf{Core drive:} Group stability through evidence-based maintenance and incremental improvement
\item \textbf{Energy from:} Incremental improvements based on evidence, maintaining excellence
\item \textbf{Archetypes:} Evidence-based policymaker, conservationist, scientific curator
\item \textbf{Career fits:} Policy analyst, museum curator, preservation specialist
\item \textbf{Challenge:} Can be overly incremental; may miss need for radical change
\end{itemize}

\textbf{[S+ O+ R- T-] The Institutional Steward}
\begin{itemize}
\item \textbf{Core drive:} Group preservation through upholding established structures and traditions
\item \textbf{Energy from:} Protecting valuable institutions, maintaining order
\item \textbf{Archetypes:} Constitutional scholar, traditional judge, institutional guardian
\item \textbf{Career fits:} Judge, archivist, constitutional lawyer, tradition keeper
\item \textbf{Challenge:} Can become rigid; may preserve past effectiveness
\end{itemize}

\textbf{[S+ O+ R+ T-] The Systematic Administrator}
\begin{itemize}
\item \textbf{Core drive:} Group stability through rational management and optimized systems
\item \textbf{Energy from:} Making existing systems run smoothly, operational excellence
\item \textbf{Archetypes:} Effective bureaucrat, operations expert, logistics master
\item \textbf{Career fits:} Operations manager, logistics coordinator, process optimizer
\item \textbf{Challenge:} Can resist innovation; may optimize for wrong metrics
\end{itemize}

\subsection{Using the Atlas}

Three functions:

\textbf{1. Diagnostic triangulation:} Archetypes producing ``jolt of recognition'' point to likely native region.

\textbf{2. Understanding Mask patterns:} Large divergence between native and performed archetype indicates strong Mask. Example: Native Gnostic Architect [S- O+ R+ T+] performing Communal Gardener [S+ O- R- T-] = full inversion, maximum overhead.

\textbf{3. Relationship dynamics:} Understanding partner's likely archetype illuminates Implicit Treaty collisions. Systematic Explorer + Communal Gardener = predictable R-axis and S-axis collision points.

\section{Extended Implicit Treaty Collision Examples}

\Cref{ch:counterfeit-self} provides detailed R-axis and S-axis collision examples. This section provides T-axis and O-axis collisions for completeness.

\subsection{T-Axis Collision: Metamorphosis vs. Homeostasis}

\textbf{Setup:}

Partner A (T+): ``I got the risky opportunity---huge growth potential but requires major upheaval! This is exactly what I've been waiting for!''

\textbf{Partner A's T+ Implicit Treaty:} ``Love means supporting growth and embracing necessary risk. Stagnation is the enemy.''

Partner B (T-): ``You want to risk our stability? Absolutely not. We've worked years to build this secure foundation.''

\textbf{Partner B's T- Implicit Treaty:} ``Love means protecting safety and stability. Unnecessary risk is the enemy.''

\textbf{The catastrophic translation:}

\textbf{Partner A hears:}
\begin{itemize}
\item My dreams don't matter
\item You want to cage me in mediocrity
\item You don't trust my capability
\item You're holding me back
\end{itemize}

\textbf{Partner B hears:}
\begin{itemize}
\item Everything we built means nothing
\item You'll gamble our security on a whim
\item You're reckless and irresponsible
\item You don't value what I've protected
\end{itemize}

\textbf{Result:}
Partner A feels caged. Partner B feels terrorized. Both experience profound betrayal.

\textbf{With power asymmetry:}
If Partner B controls resources or threatens exit: Partner A suppresses T+ ambition, performs T- contentment, hides ``reckless'' dreams.

\textbf{Outcome:} T+ native wearing counterfeit T- Mask. Predictable: Victory Trap (achieving Masked stability goals feels hollow), burnout from suppressed drive, eventual explosive ``mid-life crisis.''

\subsection{O-Axis Collision: Design vs. Emergence}

\textbf{Setup:}

Partner A (O+): ``I've optimized our Italy vacation! Complete itinerary spreadsheet with reservations, tickets, schedules, backup plans...''

\textbf{Partner A's O+ Implicit Treaty:} ``Love means planning carefully to maximize shared experience. Spontaneity without structure leads to suboptimal outcomes.''

Partner B (O-): ``A spreadsheet? You've turned our vacation into a military operation. What about wandering, discovering, being present?''

\textbf{Partner B's O- Implicit Treaty:} ``Love means being present together and letting experiences unfold naturally. Over-planning destroys magic.''

\textbf{The catastrophic translation:}

\textbf{Partner A hears:}
\begin{itemize}
\item My thoughtful gift (planning effort) is rejected
\item You're calling my care ``tyranny''
\item You don't appreciate my work to optimize our experience
\item You value chaos over my contribution
\end{itemize}

\textbf{Partner B hears:}
\begin{itemize}
\item You don't trust our connection to guide us
\item You need to control everything
\item You value efficiency over presence
\item You're turning our relationship into a project
\end{itemize}

\textbf{Result:}
Partner A feels planning is unappreciated, experienced as ``controlling.'' Partner B feels smothered by structure.

\textbf{With power asymmetry:}
If Partner B threatens exit: Partner A suppresses O+ planning, performs O- spontaneity, experiences constant anxiety from lack of structure.

\textbf{Outcome:} O+ native wearing counterfeit O- Mask. Predictable: Chronic anxiety (chaos), decision difficulty (trained to suppress planning), resentment.

\section{Cognitive Profile Deep-Dives}

\Cref{ch:counterfeit-self} provides brief profiles. This section extends the analysis.

\subsection{The Empathizer: Extended Profile}

\textbf{Approximate pSORT:} [S+ O- R- T-] region (Communal Gardener / Loyal Traditionalist)

\textbf{Core optimization:} Maintain social cohesion of group (Heart function at civilizational scale)

\textbf{Cognitive strengths:}
\begin{itemize}
\item \textbf{Social intuition:} Reads emotional states, group dynamics, relational tensions without explicit analysis
\item \textbf{Empathic resonance:} ``Feels into'' others' experience, facilitating connection and care
\item \textbf{Narrative coherence:} Understands through story, meaning, emotional arc rather than formal logic
\item \textbf{Emergent coordination:} Navigates complex social situations through intuition, not explicit rules
\item \textbf{Cultural preservation:} Maintains traditions, rituals, shared meanings binding communities
\end{itemize}

\textbf{Cognitive vulnerabilities:}
\begin{itemize}
\item \textbf{Gnostic resistance:} Analytical truth-seeking can feel ``cold'' or threatening
\item \textbf{Conflict avoidance:} May prioritize harmony over necessary truth-telling
\item \textbf{Change resistance:} Stability orientation can resist needed adaptation
\item \textbf{Boundary porosity:} High communion orientation makes maintaining personal boundaries difficult
\end{itemize}

\textbf{Evolutionary basis:} Adaptive for maintaining group cohesion---critical function throughout evolutionary history. Social harmony often more critical than analytical precision.

\textbf{Cultural fit:}
\begin{itemize}
\item \textbf{Hospice cultures} (T-, R-, S+): Natural fit. Strengths valued and rewarded.
\item \textbf{Foundry cultures} (T+, R+, S-): Poor fit without Mask. Must adopt counterfeit Systemizer traits.
\end{itemize}

\textbf{Mask pattern:} Empathizer forced into Systemizer-dominant environment (cutthroat sales, purely analytical engineering) must wear counterfeit [S- R+ T+] Mask. Feels isolating, soulless, draining. Result: burnout, loss of meaning, physical illness.

\textbf{Integrated Empathizer:} High $\Omegap$ Empathizer in aligned environment (community care, teaching, hospitality) with developed capacity to deploy R+ analysis when needed (via Skeleton) while maintaining native R- strength. Governs S+/O-/R-/T- core intelligently, doesn't suppress it.

\subsection{The Systemizer: Extended Profile}

\textbf{Approximate pSORT:} [S- O+ R+ T+] region (Gnostic Architect / Systematic Explorer)

\textbf{Universal physics:} ANY intelligent system optimizing for individual agency (S-), designed order (O+), truth-seeking (R+), and growth (T+) exhibits Systemizer patterns. This is solution to Trinity of Tensions, not human-specific trait. AGI with these targets would display analogous profile.

\textbf{Human implementation:} Correlates with left-hemisphere dominance (\Cref{ch:biological}). Often clinically labeled as \textbf{Autism Spectrum (ASD)} or \textbf{ADHD} (hyperfocus subtype) in cultures pathologizing this orientation.

\textbf{Core optimization:} Understand and optimize systems for improved performance (Head function at civilizational scale)

\textbf{Cognitive strengths:}
\begin{itemize}
\item \textbf{Systematic analysis:} Decomposes complex systems into components, identifies failure modes
\item \textbf{Truth-seeking:} Prioritizes accuracy over social comfort; states uncomfortable truths
\item \textbf{Pattern recognition:} Identifies structural similarities across domains; builds generalizable models
\item \textbf{Design optimization:} Sees how systems could be improved; instinctive engineering orientation
\item \textbf{Growth orientation:} Constantly learning, building, expanding capability
\item \textbf{Principled consistency:} Applies same logical standards universally, resistant to special pleading
\end{itemize}

\textbf{Cognitive vulnerabilities:}
\begin{itemize}
\item \textbf{Social intuition deficit:} Doesn't naturally read emotional subtext or group dynamics
\item \textbf{Empathic translation:} Can provide R+ analysis when R- validation needed (collision pattern)
\item \textbf{Obsessive focus:} Can hyperfocus on systems at expense of relationships or self-care
\item \textbf{Change insistence:} T+ drive to improve experienced as ``never satisfied''
\item \textbf{Boundary rigidity:} High S- makes communion and interdependence challenging
\end{itemize}

\textbf{Evolutionary basis:} Adaptive for tool-making, system optimization, innovation---critical for civilizational advancement but not necessary for every tribe member. Lower base rate than Empathizer.

\textbf{Cultural fit:}
\begin{itemize}
\item \textbf{Foundry cultures} (T+, R+, S-): Natural fit. Strengths valued. Silicon Valley, elite research, rationalist communities.
\item \textbf{Hospice cultures} (T-, R-, S+): Poor fit without Mask. Must adopt counterfeit Empathizer traits.
\end{itemize}

\textbf{The ``AuDHD'' experience:} Gnostic Architect forced to pretend it's Loyal Traditionalist. Must:
\begin{itemize}
\item Suppress analytical impulses (``you're being too critical'')
\item Perform social intuition they lack (``read the room'')
\item Feign interest in small talk and social rituals
\item Hide T+ drive to fix ``broken'' systems (``stop trying to change everything'')
\item Pretend S- autonomy needs don't exist (``why do you always need to be alone?'')
\end{itemize}

Resulting low $\Omegap$ and chronic burnout are predictable physics. Mask overhead (suppression + performance + translation + monitoring) consumes 40-70\% of cognitive capacity.

\textbf{Clinical vs. framework perspective:}

\textbf{Clinical} (ASD, ADHD): Treats Systemizer profile as pathology requiring treatment. Focus on ``deficits'' (social skills training, emotional regulation).

\textbf{Framework:} Systemizer profile is valid solution to Four Axiomatic Dilemmas. ``Pathology'' is environmental mismatch (Systemizer in Hospice culture) requiring Mask. Treatment target: increase environmental compatibility OR build Internal Polity enabling sustainable operation in mismatched contexts.

\textbf{Integrated Systemizer:} High $\Omegap$ Systemizer in aligned environment (research, engineering, rationalist community) with developed capacity to deploy R- communion and S+ connection when needed (via Skeleton) while maintaining native S-/O+/R+/T+ core. Governs analytical precision with wisdom, doesn't suppress it.

\section{Neurodiversity as Axiological Physics: Extended Analysis}

\Cref{ch:counterfeit-self} introduces neurodiversity as cultural-scale Implicit Treaty collision. Extended analysis follows.

\subsection{The Pathologization Pattern}

When entire culture runs dominant Implicit Treaty optimized for one cognitive profile, individuals with incompatible profiles face systematic Mask pressure. Medical model's ``disorder'' diagnosis often reflects cultural mismatch rather than inherent dysfunction.

\textbf{Key insight:} Many ``disorders'' are high-overhead Mask states, not biological pathologies.

\textbf{Testable predictions:}

\begin{itemize}
\item \textbf{Cross-cultural variance:} ASD/ADHD prevalence should correlate with culture's axiological distance from Systemizer optimal. Prediction: Lower diagnosis in cultures valuing individual competence, analytical precision, innovation.

\item \textbf{Gender ratios:} ASD diagnosed 4:1 male:female. Framework hypothesis: Not biological male vulnerability but differential Mask pressure. Males face less pressure to perform Empathizer traits, so native Systemizers remain unmasked longer. Females with identical profiles learn Empathizer Mask earlier, masking ``symptoms'' until Mask failure (often adulthood).

\item \textbf{Late diagnosis:} Framework predicts: Adult-diagnosed ASD/ADHD are high-functioning Systemizers whose Masks failed after decades. Precipitating events: Mask-demanding job/relationship ended, burnout made performance unsustainable, life complexity exceeded Mask capacity.

\item \textbf{Burnout correlation:} Neurodivergent individuals in Hospice cultures should show significantly higher burnout than neurotypical peers, controlling for hours worked. Mechanism: Mask overhead. Falsification: If burnout rates equivalent after controlling for work hours, Mask theory insufficient.

\item \textbf{``Camouflaging'' research:} Devon Price's \textit{Unmasking Autism} (2022) provides extensive phenomenological documentation of the costs of autistic masking---the exhaustion, burnout, and identity fragmentation from sustained performance of neurotypical behavior. Price's empirical observations of masking's toll align precisely with this framework's thermodynamic prediction: running counterfeit pSORT on incompatible substrate produces measurable $\Omegap$ loss. Existing research on ``autistic camouflaging'' (suppressing autistic traits to appear neurotypical) should map onto Mask framework. Prediction: High camouflaging scores correlate with low $\Omegap$, high burnout, increased mental health issues.
\end{itemize}

\subsection{Treatment Implications}

\textbf{Medical Model approach:}
\begin{itemize}
\item Diagnosis: Individual has disorder
\item Treatment: Train to approximate neurotypical behavior (social skills training, behavioral therapy)
\item Goal: Reduce ``symptoms'' (increase masking effectiveness)
\end{itemize}

\textbf{Axiological Mismatch Model approach:}
\begin{itemize}
\item Diagnosis: Native pSORT mismatched with environmental demands, generating high-overhead Mask
\item Treatment options:
  \begin{itemize}
  \item \textbf{Environmental modification:} Shift to contexts aligned with native profile (technical roles for Systemizers, care roles for Empathizers)
  \item \textbf{Strategic Unmasking:} Reduce Mask intensity in Safe Zones, build Internal Polity as replacement
  \item \textbf{Skills as tools, not Masks:} Develop O+ structured social scripts or R- empathic attunement as conscious tools deployed strategically, not counterfeit core identity
  \end{itemize}
\item Goal: Increase $\Omegap$ (reduce internal conflict), not increase Mask effectiveness
\end{itemize}

\textbf{Key distinction---social skills as tools vs. Mask:}
\begin{itemize}
\item \textbf{Tool:} Systemizer learns structured social scripts as deliberate techniques deployed in specific contexts. Conscious, strategic, doesn't suppress native identity. Low overhead. Example: ``I will ask three questions about the other person before discussing my interests.''
\item \textbf{Mask:} Systemizer performs complete Empathizer persona---pretending to intuitively read emotional subtext, feigning interest in small talk, suppressing all analytical impulses. Unconscious, pervasive, suppresses native identity. High overhead.
\end{itemize}

Tool-based approach increases capability without reducing $\Omegap$. Mask-based approach increases social ``passing'' while reducing $\Omegap$ and causing burnout.

\subsection{Research Directions}

To validate or falsify axiological mismatch model:

\textbf{1. pSORT-Environment Fit Studies:}
Measure native pSORT (once validated instrument exists), measure environmental axiological demands, predict $\Omegap$ and burnout from mismatch magnitude. Falsification: If pSORT-environment mismatch has no predictive power beyond traditional predictors.

\textbf{2. Longitudinal Mask Tracking:}
Follow neurodivergent individuals through different environmental contexts (Mask-demanding vs. Mask-optional). Framework predicts: $\Omegap$ and dysfunction should vary with environmental match, not remain constant. Falsification: If dysfunction context-independent.

\textbf{3. Cross-Cultural Neurodivergence Studies:}
Compare ASD/ADHD prevalence across cultures with different axiological configurations. Framework predicts: Cultures more aligned with Systemizer profile should show lower diagnosis rates and lower dysfunction for same trait profiles. Falsification: If prevalence and dysfunction culturally invariant.

\textbf{4. Unmasking Intervention Trials:}
Randomized controlled trials comparing:
\begin{itemize}
\item Group A: Traditional social skills training (increase Mask effectiveness)
\item Group B: Unmasking protocol (reduce Mask, build Internal Polity)
\item Group C: Control
\end{itemize}
Measure: $\Omegap$, burnout, social functioning, quality of life. Framework predicts: Group B shows higher $\Omegap$ and lower burnout than Group A despite possibly lower social ``passing.'' Falsification: If Group A outcomes superior or equivalent.

\section{Novel Contributions and Falsification Roadmap}

\subsection{What Framework Adds}

\textbf{Thermodynamic grounding:} pSORT dimensions derived from universal physical constraints (Four Axiomatic Dilemmas) any negentropic agent must navigate, not from empirical clustering or clinical observation. Provides claimed universality (should apply to any intelligent system), non-arbitrariness (necessary consequences of physics), and predictive potential (stable vs. pathological configurations from first principles).

\textbf{Holographic integration:} Same physics governing civilizations governs individual psyches. $\Omegap \to \Omega$ aggregation is mechanistic claim that personal Coherence is atomic unit of civilizational Coherence. Connects personal integration to civilizational Re-Founding (\Cref{ch:great-work}).

\textbf{AI alignment connection:} No prior personality framework connects personal integration to AGI alignment. Framework proposes: Goal misalignment in humans (Mask) parallels AI systems (mesa-optimization, reward hacking). Human Unmasking training ground for understanding AGI alignment challenge. Integrated Humans necessary infrastructure for navigating intelligence explosion safely.

\textbf{Falsifiable architecture:} Complete top-to-bottom framework---from physics to diagnostics to engineering blueprints---presented as falsifiable hypotheses, inviting scientific approach to human flourishing.

\subsection{Individual-Level Predictions}

\begin{itemize}
\item \textbf{pSORT Predictive Validity:} Hypothesized native signatures should predict energy patterns. Activities aligned with diagnosed native pSORT should measurably energize; misaligned should drain. \textit{Falsification:} No correlation between diagnosed pSORT and energy patterns after controlling for confounds.

\item \textbf{Neural Correlates:} Native pSORT and Mask signatures should show distinct neural activation patterns. \textit{Falsification:} fMRI studies find no distinguishable patterns.

\item \textbf{Longitudinal $\Omegap$ Increases:} Unmasking protocol should produce replicable $\Omegap$ increases in longitudinal studies. \textit{Falsification:} RCTs show no significant increase vs. controls.

\item \textbf{Big Five Correlations:} pSORT axes should correlate with Big Five as hypothesized (T↔Openness, O↔Conscientiousness, S↔Extraversion, Neuroticism↔low $\Omegap$). \textit{Falsification:} Large-scale studies find correlation coefficients below 0.3.
\end{itemize}

\subsection{Population-Level Predictions}

\begin{itemize}
\item \textbf{Mask Prevalence:} Cross-cultural studies should find higher Mask prevalence in high-social-conformity cultures. \textit{Falsification:} No correlation between cultural conformity metrics and self-reported Masking.

\item \textbf{Neurodivergence Patterns:} Mask prevalence significantly higher among neurodivergent populations in cultures axiologically misaligned with typical Systemizer pSORT. \textit{Falsification:} No difference in Mask rates between neurodivergent and neurotypical populations.

\item \textbf{Mask-Burnout Link:} Divergence between native pSORT and job-required pSORT should predict burnout better than hours worked. \textit{Falsification:} Regression shows pSORT divergence adds no explanatory power beyond traditional predictors.

\item \textbf{Mid-Life Crisis:} ``Victory Trap'' experiences (achieving Mask-driven goals feeling hollow) should cluster in mid-life among high-achievers. \textit{Falsification:} No age-related pattern or correlation with Mask strength.
\end{itemize}

\subsection{N-of-1 Validation}

Central testable claim of your personal experiment:

\begin{quote}
\textbf{Hypothesis:} By using diagnostic lenses to reduce divergence between my behavior and native pSORT, my personal Coherence ($\Omegap$) will measurably increase over 3-6 months.
\end{quote}

\textbf{Evidence of validation:}
After 3-6 months honest engagement: measurably more energy, greater clarity, reduced internal conflict, greater action capacity. Framework provided useful explanatory power.

\textbf{Evidence of falsification:}
After 3-6 months honest effort: no discernible improvement. Energy levels remain low or decrease. Reducing divergence provides no energetic benefit. Mask and pSORT concepts fail to provide new, useful explanatory power.

If falsified in your experiment: Either framework doesn't apply to your psychology, or initial pSORT diagnosis incorrect. Gnostic response: Return to diagnosis with new data, form new hypothesis.

\subsection{Validation Pathways}

V1.0 framework---invitation to distributed validation. Your N-of-1 experiment contributes to larger project, whether results positive or negative.

\textbf{Three validation paths:}
\begin{enumerate}
\item \textbf{Phenomenological:} Does framework resonate with lived experience? Provide explanatory power for observed patterns?
\item \textbf{Instrumental:} Does applying protocol produce measurable outcomes (increased $\Omegap$, sustained T+ project progress)?
\item \textbf{Empirical:} Do testable predictions (Big Five correlations, neural patterns, population prevalence) hold under systematic investigation?
\end{enumerate}

Negative results at any level are valuable data. Framework robustness depends on honest reporting of successes and failures.

Successful execution of this research program required to move \Cref{part:integrated-human} from Tier 2 theoretical synthesis to validated, Tier 1 science. Until then, it remains lenses and framework for your Gnostic experiment.

\textbf{The work begins with you.}
