
\vspace{0.5em}

\epistemicmixed{Mixed Confidence (Tier 1-3)}{
Core mechanism—Four Axiomatic Dilemmas apply to individual psyches with same necessity they apply to civilizations—is Tier 1 physics. pSORT framework and Personal Iron Law are direct applications. Diagnostic methods (self-assessment, archetypal recognition, Mask identification) are Tier 2-3 structured introspection tools. Appropriate epistemology: N-of-1 validation. Does this understanding reduce internal conflict? That's the test. See Appendix D for detailed methodology.}

\section{The Constraint Is Physics}

You finished Part IV with a blueprint for civilizational Re-Founding. You understand the physics, the engineering principles, the governance architecture. You know what needs to be built.

\vspace{0.5em}

Part V solves a thermodynamic constraint that makes civilizational ambitions impossible without personal integration.

\vspace{0.5em}

\textbf{The Personal Iron Law of Coherence:}

Personal Coherence below critical threshold (\Omegap $< 0.3$) makes sustained, constructive Action (\Alphap+) impossible.

\vspace{1em}

The mechanism is pure energy accounting. Low \Omegap{} is a state of internal civil war. Energy required to manage the conflict—to suppress your native impulses, to perform a counterfeit persona, to translate between incompatible axiological systems—is consumed internally. There is no surplus energy left to project into the world as constructive action.

\vspace{0.5em}

\textbf{Implication:} Personal Aliveness is the atomic unit of civilizational Aliveness. Every person who increases their \Omegap{} is a direct contribution to the coherence of the whole. This is civilizational engineering at the atomic scale.

\section{The Holographic Application}

Part V demonstrates framework universality by applying it at personal scale. The same Four Axiomatic Dilemmas that govern civilizations govern individual psyches. If framework describes universal physics of telic systems, it must apply to you.

The foundation is identical: any telic system—virus, organism, civilization, individual psyche, AGI—faces the same physical constraints. Thermodynamic dilemmas, boundary problems, information trade-offs, control coordination. The physics is scale-invariant.

What varies is the substrate and the scale, not the constraint. The framework's power comes from its universality. Personal application follows directly from applying universal physics to a different scale.

\vspace{0.5em}

\section{The Path Forward}

Part V is structured as a single, coherent, three-act journey into the physics of the self.

\begin{description}
    \item[Chapter 18: The Physics of the Self] The Four Axiomatic Dilemmas apply to individual psyche with same necessity as civilizations. The Personal Iron Law proves why low \Omegap{} makes sustained \Alphap+ impossible.

    \item[Chapter 19: The Counterfeit Self] The primary pathology of low \Omegap: the Mask. Understand its origin in Implicit Treaty collisions—when the native pSORT confronts incompatible environmental demands. Formalizing Implicit Treaties reveals relationship conflicts as architectural problems.

    \item[Chapter 20: The Integrated Human] The engineering blueprint: 3-Layer Internal Polity that replaces the Mask. Diagnostic lenses for making the Mask's structure legible. The Shared Skeleton for scaling integration from individual to dyad. Your N-of-1 experiment with explicit falsification criteria.
\end{description}

\vspace{0.5em}

The journey's end is understanding. The proof of the holographic principle is that you can use the same physics that built this framework to understand yourself.

The Re-Founding begins within.
